|

### 2023-06-21 presentations/2023-NECTAR-coimbra.qmd
Updated NECTAR presentation

---
title: |
    | Profiling 
    | the 15-min *pedshed* 
    | from the perspective of 
    | network attributes
author: "**Antonio Páez**, Belén Martín, Josh Arbess, James R. Dunn"
format: revealjs
execute: 
  cache: true
---

@@

load(file = glue(here(), "/data-raw/amenities_walkshed.rda"))
load(file = glue(here(), "/data-raw/hamilton_net.rda"))
load(file = glue(here(), "/data-raw/urban_hammer_da.rda"))
data("urban_types")
load(file = glue(here(), "/data-raw/walksheds.rda"))
load(file = glue(here(), "/data-raw/walksheds_net_attributes.rda"))
```

```{r clean-net-attributes-data}
#| include: false
d <- walksheds_net_attributes |>
                  drop_na()
```

@@

A pedestrian shed $\equiv$ a 15-min isochrone.

@@

global <- walksheds |>
  mutate(area = st_area(geometry),
         area = units::set_units(area, km^2) |> 
           drop_units()) |>
  st_drop_geometry() |>
  select(-c(GeoUID, Type)) |>
  skim()

by_type <- walksheds |>
  mutate(area = st_area(geometry),
         area = units::set_units(area, km^2) |> 
           drop_units()) |>
  st_drop_geometry() |>
  select(-c(GeoUID)) |>
  group_by(Type) |>
  skim()

@@

@@

@@

- Transitivity (probability that the friends of my friends are my friends: minimal circuits).
- Motifs (recurrence of graph elements: repetition of patterns).
- Diameter (longest shortest path in network: how distant are the two most distant nodes).
- Radius (shortest shortest path in network: how close are the two closest nodes).
- Girth (length of shortest circle: shortest path for a node to reach itself).
- Global efficiency (harmonic mean of distances between all nodes: how easy it is on average to reach others).
- Edge density (ratio of actual to maximum edges: how connected is the network).
- Edge connectivity (How many edges need to be removed to eliminate all paths between two nodes).
- Mean distance (mean of length of all shortest paths).

@@

# This is Tree 3 with seed 3 (46125) in 05-Walkshed-Profiles.Rmd
seed3 <- 46125
set.seed(seed3)
tree3 <- evtree(Type ~ ., 
                data = d |> select(-GeoUID))
```

@@

ggparty(tree3) +
  geom_edge() +
  geom_edge_label(aes(label = str_replace_all(breaks_label, 
                                          "0.0207311636201516", 
                                          "0.021"))) +
  geom_node_splitvar(aes(label = ifelse(kids == 0,
                                        glue(" n = {nodesize}"), 
                                        splitvar)),
                     nudge_x = 0.0,
                     ids = c(1, 2, 5)) +
  geom_node_splitvar(aes(label = ifelse(kids == 0,
                                        glue(" n = {nodesize}"), 
                                        splitvar)),
                     nudge_x = 0.02,
                     ids = c(3, 4, 6, 7)) +
  geom_node_plot(gglist = list(geom_bar(aes(x = "", 
                                            fill = Type),
                                        position = position_fill()),
                               xlab("Type"),
                               theme_minimal()),
                 # draw only one label for each axis
                 shared_axis_labels = TRUE,
                 # draw line between tree and legend
                 legend_separator = FALSE)
```

@@

set.seed(69717)

leaf_1_plot <- st_intersection(hamilton_net$edges,
             walksheds |> 
  filter(GeoUID == (walksheds_net_attributes |> 
                      filter(transitivity < 0.0207311636201516,
                             motifs < 378, 
                             Type == "Suburban") |> 
                      slice_sample(n=1) |> 
                      pull(GeoUID)))) |>
  ggplot() +
  geom_sf() +
  ggtitle("Transitivity < 0.021, motifs < 378") +
  theme_void()
```

@@

leaf_3_plot <- st_intersection(hamilton_net$edges,
             walksheds |> 
  filter(GeoUID == (walksheds_net_attributes |> 
                      filter(transitivity >= 0.0207311636201516,
                             motifs < 1024, 
                             Type == "Suburban") |> 
                      slice_sample(n=1) |> 
                      pull(GeoUID)))) |>
  ggplot() +
  geom_sf() +
  ggtitle("Transitivity >= 0.021, motifs < 1,024") +
  theme_void()
```

@@

set.seed(44785)

leaf_2_plot <- st_intersection(hamilton_net$edges,
             walksheds |> 
  filter(GeoUID == (walksheds_net_attributes |> 
                      filter(transitivity < 0.0207311636201516,
                             motifs >= 378, 
                             Type == "Urban") |> 
                      slice_sample(n=1) |> 
                      pull(GeoUID)))) |>
  ggplot() +
  geom_sf() +
  ggtitle("Transitivity < 0.021, motifs >= 378") +
  theme_void()
```

@@

set.seed(81461)

leaf_4_plot <- st_intersection(hamilton_net$edges,
             walksheds |> 
  filter(GeoUID == (walksheds_net_attributes |> 
                      filter(transitivity >= 0.0207311636201516,
                             motifs >= 1024, 
                             Type == "Urban") |> 
                      slice_sample(n=1) |> 
                      pull(GeoUID)))) |>
  ggplot() +
  geom_sf() +
  ggtitle("Transitivity >= 0.021, motifs >= 1,024") +
  theme_void()
```

@@

d <- d |>
  mutate(leaf = case_when(transitivity < 0.0207311636201516 & motifs < 378 ~ "Leaf 1",
                          transitivity < 0.0207311636201516 & motifs >= 378 ~ "Leaf 2",
                          transitivity >= 0.0207311636201516 & motifs < 1024 ~ "Leaf 3",
                          transitivity >= 0.0207311636201516 & motifs >= 1024 ~ "Leaf 4"),
         leaf = factor(leaf))
```

@@

d <- d |>
  left_join(walksheds |>
              mutate(area = st_area(geometry),
                     area = units::set_units(area, km^2) |> 
                       drop_units()) |>
              st_drop_geometry(),
            by = c("GeoUID", "Type")) 
```

@@

d |>
  select(area, leaf) |>
  group_by(leaf) |>
  skim() |>
  mutate(across(numeric.mean:numeric.p100, 
                      ~round(.x, 2)),
         Type = c("Typically suburban", "Typically urban", "Typically suburban", "Typically urban")) |>
  select(leaf, Type, starts_with("numeric")) |>
  rename_with(~ str_replace(.x, "numeric.", ""), starts_with("numeric")) |>
  gt() |>
  cols_width(everything() ~ px(100)) |>
  tab_options(table.font.size = 30)
```

@@

```{r estimate-model-area}
#| include: false
mdl <- d |>
  mutate(area = log(area),
         transitivity_2 = transitivity^2,
         motifs = motifs/1000,
         motifs_2 = (motifs)^2) |>
  lm(area ~ transitivity + transitivity_2 + motifs + motifs_2, data = _)
```

@@

mdl_tidy <- mdl |>
  tidy() |>
  mutate(term = c("intercept", "transitivity", "transitivity^2", "motifs", "motifs^2"))
```

@@

b2 <- mdl_tidy |> filter(term == "transitivity") |> pull(estimate)
b3 <- mdl_tidy |> filter(term == "transitivity^2") |> pull(estimate)
f_transitivity <- function(x) exp(b1 + b2 * x +  b3 * x^2)

effect_transitivity <- ggplot() +
  geom_function(fun = f_transitivity,
                xlim = c(0, 0.12)) +
    ylim(c(0.6, 1.6)) +
  ggtitle("Expected area as a function of transitivity") + 
  ylab("area (sq.km)") +
  xlab("transitivity") +
  theme_minimal()
```

@@

b1 <- mdl_tidy |> filter(term == "intercept") |> pull(estimate)
b4 <- mdl_tidy |> filter(term == "motifs") |> pull(estimate)
b5 <- mdl_tidy |> filter(term == "motifs^2") |> pull(estimate)
f_motifs <- function(x) exp(b1 + b4 * x +  b5 * x^2)

effect_motifs <- ggplot() +
  geom_function(fun = f_motifs,
                xlim = c(0, 6.824)) +
  ylim(c(0.6, 1.6)) +
  ggtitle("Expected area as a function of motifs (in 1,000s)") + 
  ylab("area (sq.km)") +
  xlab("motifs (1,000s)") +
  theme_minimal()
```

@@

effect_transitivity + effect_motifs
```

@@

- pedshed area varies greatly within Hamilton.
- The size of pedsheds varies significantly with the attributes of the network.
- The effect is non-linear.
- Results suggest an optimal combination of attributes to maximize pedshed areas.

@@

- Expand the analysis to include the environment, i.e., the opportunity landscape.
- Try other cities.

>>>>

---
title: |
    | Reality or wishful thinking:
    | Network attributes of 15-minute neighbourhoods
author: "**Antonio Páez**, Belén Martín, Josh Arbess, James R. Dunn"
format: revealjs
execute: 
  cache: true
---

@@

load(file = glue(here(), "/data-raw/amenities_walkshed.rda"))
load(file = glue(here(), "/data-raw/hamilton_net.rda"))
load(file = glue(here(), "/data-raw/urban_hammer_da.rda"))
data("urban_types")
load(file = glue(here(), "/data-raw/walksheds_da.rda"))
load(file = glue(here(), "/data-raw/walksheds_da_net_vars.rda"))
hamilton_edges <- hamilton_net$edges
```

```{r clean-net-attributes-data}
#| include: false
d <- walksheds_da_net_vars |>
  drop_na() |>
  mutate(normalized_motifs_3 = motifs_3/n_edges,
         normalized_motifs_4 = motifs_4/n_edges)
```

@@

A 15-min pedestrian shed $\equiv$ a 15-min walking isochrone.

@@

global <- walksheds_da |>
  mutate(area = st_area(geometry),
         area = units::set_units(area, km^2) |> 
           drop_units()) |>
  st_drop_geometry() |>
  select(-c(GeoUID, Type)) |>
  skim()

by_type <- walksheds_da |>
  mutate(area = st_area(geometry),
         area = units::set_units(area, km^2) |> 
           drop_units()) |>
  st_drop_geometry() |>
  select(-c(GeoUID)) |>
  group_by(Type) |>
  skim()

@@

```{r}
#| include: false

#seed <- sample.int(29873, 1)
#seed
set.seed(20512)

sample_walkshed <- walksheds_da |>
  filter(Type == "Urban") |>
  slice_sample(n = 1)
#  filter(GeoUID == "35250102")

# Find the intersection between the edges and the walksheds. The edges will be truncated, but the indices can be used to retrieve the full edges.
wsnet <- hamilton_edges |>
  st_filter(
    sample_walkshed,
    .predicate = st_intersects)

# Extract the full edges that were found to intersect the walkshed
wsnet <- hamilton_edges |> 
  filter(edge_index %in% wsnet$edge_index)

sample_da <- urban_hammer_da |>
  filter(GeoUID == sample_walkshed$GeoUID)

urban_walkshed <- ggplot() +
  geom_sf(data = sample_walkshed) +
  geom_sf(data = sample_da,
          fill = NA,
          color = "#619CFF",
          linewidth = 4) +
  geom_sf(data = wsnet) +
  theme_void()

```{r}
#| include: false

#seed <- sample.int(29873, 1)
#seed
#137

set.seed(137)

sample_walkshed <- walksheds_da |>
  filter(Type == "Suburban") |>
  slice_sample(n = 1)
#  filter(GeoUID == "35250102")

# Find the intersection between the edges and the walksheds. The edges will be truncated, but the indices can be used to retrieve the full edges.
wsnet <- hamilton_edges |>
  st_filter(
    sample_walkshed,
    .predicate = st_intersects)

# Extract the full edges that were found to intersect the walkshed
wsnet <- hamilton_edges |> 
  filter(edge_index %in% wsnet$edge_index)

sample_da <- urban_hammer_da |>
  filter(GeoUID == sample_walkshed$GeoUID)

suburban_walkshed <- ggplot() +
  geom_sf(data = sample_walkshed) +
  geom_sf(data = sample_da,
          fill = NA,
          color = "#F8766D",
          linewidth = 4) +
  coord_sf(xlim = c(-79.94532, -79.91076),
           ylim = c(43.25742, 43.27836)) +
  geom_sf(data = wsnet) +
  theme_void()

@@

urban_walkshed + suburban_walkshed
```

@@

- Edge density (ratio of actual to maximum edges: how connected is the network).
- Motifs (recurrence of graph elements: repetition of patterns).
- Diameter (longest shortest path in network: how distant are the two most distant nodes).
- Radius (shortest shortest path in network: how close are the two closest nodes).
- Girth (length of shortest circle: shortest path for a node to reach itself).
- Global efficiency (harmonic mean of distances between all nodes: how easy it is on average to reach others).
- Transitivity (probability that the friends of my friends are my friends: minimal circuits).
- Edge connectivity (How many edges need to be removed to eliminate all paths between two nodes).
- Mean distance (mean of length of all shortest paths).

@@

#seed3 <- sample.int(1000000, 1)
#seed3 <- 853251
#seed3
set.seed(853251)
tree3 <- evtree(Type ~ ., 
                data = d |> select(-c(GeoUID, motifs_3, motifs_4, n_edges, n_nodes)))
#plot(tree3)
```

@@

ggparty(tree3) +
  geom_edge() +
  geom_edge_label(aes(label = str_replace_all(breaks_label, 
                                              c("0.00591859212548868" = "0.006",
                                                "0.86734693877551" = "0.867",
                                                "0.967741935483871" = "0.968")))) +
  geom_node_splitvar(aes(label = ifelse(kids == 0,
                                        glue(" n = {nodesize}"), 
                                        splitvar)),
                     nudge_x = 0.0,
                     ids = c(1, 2, 5)) +
  geom_node_splitvar(aes(label = ifelse(kids == 0,
                                        glue(" n = {nodesize}"), 
                                        splitvar)),
                     nudge_x = 0.02,
                     ids = c(3, 4, 6, 7)) +
  geom_node_plot(gglist = list(geom_bar(aes(x = "", 
                                            fill = Type),
                                        position = position_fill()),
                               xlab("Type"),
                               theme_minimal()),
                 # draw only one label for each axis
                 shared_axis_labels = TRUE,
                 # draw line between tree and legend
                 legend_separator = FALSE)
```

@@

set.seed(289919)

leaf_1_plot <- st_intersection(hamilton_net$edges,
             walksheds_da |> 
  filter(GeoUID == (walksheds_da_net_vars |> 
                      mutate(normalized_motifs_3 = motifs_3/n_edges) |>
                      filter(edge_density < 0.00591859212548868,
                             normalized_motifs_3 < 0.86734693877551, 
                             Type == "Suburban") |> 
                      slice_sample(n=1) |> 
                      pull(GeoUID))))

leaf_1_plot <- hamilton_net$edges |>
  filter(edge_index %in% leaf_1_plot$edge_index)

leaf_1_plot <- leaf_1_plot |>
  ggplot() +
  geom_sf() +
  ggtitle("edge density < 0.006, motifs < 0.867") +
  theme_void()
```

@@

leaf_3_plot <- st_intersection(hamilton_net$edges,
             walksheds_da |> 
  filter(GeoUID == (walksheds_da_net_vars |> 
                      mutate(normalized_motifs_3 = motifs_3/n_edges) |>
                      filter(edge_density >= 0.00591859212548868,
                             normalized_motifs_3 < 0.967741935483871, 
                             Type == "Suburban") |> 
                      slice_sample(n=1) |> 
                      pull(GeoUID))))

leaf_3_plot <- hamilton_net$edges |>
  filter(edge_index %in% leaf_3_plot$edge_index)

leaf_3_plot <- leaf_3_plot |>
  ggplot() +
  geom_sf() +
  ggtitle("edge density < 0.006, motifs >= 0.968") +
  theme_void()
```

@@

set.seed(92348)

leaf_2_plot <- st_intersection(hamilton_net$edges,
             walksheds_da |> 
  filter(GeoUID == (walksheds_da_net_vars |> 
                      mutate(normalized_motifs_3 = motifs_3/n_edges) |>
                      filter(edge_density < 0.00591859212548868,
                             normalized_motifs_3 >= 0.86734693877551, 
                             Type == "Urban") |> 
                      slice_sample(n=1) |> 
                      pull(GeoUID))))

leaf_2_plot <- hamilton_net$edges |>
  filter(edge_index %in% leaf_2_plot$edge_index)

leaf_2_plot <- leaf_2_plot |>
  ggplot() +
  geom_sf() +
  ggtitle("edge density < 0.006, motifs >= 0.867") +
  theme_void()
```

@@

set.seed(392047)

leaf_4_plot <- st_intersection(hamilton_net$edges,
             walksheds_da |> 
  filter(GeoUID == (walksheds_da_net_vars |> 
                      mutate(normalized_motifs_3 = motifs_3/n_edges) |>
                      filter(edge_density >= 0.00591859212548868,
                             normalized_motifs_3 >= 0.967741935483871, 
                             Type == "Urban") |> 
                      slice_sample(n=1) |> 
                      pull(GeoUID))))

leaf_4_plot <- hamilton_net$edges |>
  filter(edge_index %in% leaf_4_plot$edge_index)

leaf_4_plot <- leaf_4_plot |>
  ggplot() +
  geom_sf() +
  ggtitle("edge density < 0.006, motifs >= 0.968") +
  theme_void()
```

@@

d <- d |>
  mutate(leaf = case_when(edge_density < 0.00591859212548868 & normalized_motifs_3 < 0.86734693877551 ~ "Leaf 1",
                          edge_density < 0.00591859212548868 & normalized_motifs_3 >= 0.86734693877551 ~ "Leaf 2",
                          edge_density >= 0.00591859212548868 & normalized_motifs_3 < 0.967741935483871 ~ "Leaf 3",
                          edge_density >= 0.00591859212548868 & normalized_motifs_3 >= 0.967741935483871 ~ "Leaf 4"),
         leaf = factor(leaf))
```

@@

d <- d |>
  left_join(walksheds_da |>
              mutate(area = st_area(geometry),
                     area = units::set_units(area, km^2) |> 
                       drop_units()) |>
              st_drop_geometry(),
            by = c("GeoUID", "Type")) 
```

@@

d |>
  select(area, leaf) |>
  group_by(leaf) |>
  skim() |>
  mutate(across(numeric.mean:numeric.p100, 
                ~round(.x, 2)),
         Type = c("Typically suburban", "Typically urban", "Typically suburban", "Typically urban"),
         n = c(d |> filter(leaf == "Leaf 1") |> nrow(),
               d |> filter(leaf == "Leaf 2") |> nrow(),
               d |> filter(leaf == "Leaf 3") |> nrow(),
               d |> filter(leaf == "Leaf 4") |> nrow())) |>
  select(leaf, n, Type, starts_with("numeric")) |>
  rename_with(~ str_replace(.x, "numeric.", ""), starts_with("numeric")) |>
  gt() |>
  cols_width(everything() ~ px(100)) |>
  tab_options(table.font.size = 30)
```

@@

```{r estimate-model-area}
#| include: false
mdl <- d |>
  mutate(area = log(area),
         edge_density_2 = edge_density^2,
         normalized_motifs_3_2 = normalized_motifs_3^2) |>
  lm(area ~ edge_density + edge_density_2 + normalized_motifs_3, data = _)
```

@@

mdl_tidy <- mdl |>
  tidy() |>
  mutate(term = c("intercept", "edge_density", "edge_density^2", "normalized_motifs_3"))
```

@@

b1 <- mdl_tidy |> filter(term == "intercept") |> pull(estimate)
b2 <- mdl_tidy |> filter(term == "edge_density") |> pull(estimate)
b3 <- mdl_tidy |> filter(term == "edge_density^2") |> pull(estimate)
f_transitivity <- function(x) exp(b1 + b2 * x +  b3 * x^2)

effect_edge_density <- ggplot() +
  geom_function(fun = f_transitivity,
                xlim = c(0, 0.4)) +
    ylim(c(0.05, 1.9)) +
  ggtitle("Expected area as a function of edge density") + 
  ylab("area (sq.km)") +
  xlab("edge density") +
  theme_minimal()
```

@@

b1 <- mdl_tidy |> filter(term == "intercept") |> pull(estimate)
b4 <- mdl_tidy |> filter(term == "normalized_motifs_3") |> pull(estimate)
f_motifs <- function(x) exp(b1 + b4 * x)

effect_motifs <- ggplot() +
  geom_function(fun = f_motifs,
                xlim = c(0.45, 1.1)) +
  ylim(c(0.05, 1.9)) +
  ggtitle("Expected area as a function of normalized motifs") + 
  ylab("area (sq.km)") +
  xlab("normalized motifs (size = 3)") +
  theme_minimal()
```

@@

effect_edge_density + effect_motifs
```

@@

- pedshed area varies greatly within Hamilton.
- The size of pedsheds varies significantly with the attributes of the network.
- Results suggest pedshed areas maximized when many connections built with repetitive patterns.

@@

- Expand the analysis to include the environment, i.e., the opportunity landscape.
- Examine other cities.
